---
title: "DSC 4824 assignment 1"
author: "Nkanyiso Ntombela 46442871"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Run the Tabu algorithm on the given matrix
0, 5, 4, 8, 7, 5
5, 0, 4, 6, 3, 2,
4, 4, 0, 5, 8, 1,
8, 6, 5, 0, 6, 4,
7, 3, 8, 6, 0, 6,
5, 2, 1, 4, 6, 0
Give the result and best tour. Run it a few times to see if the best tour changes.

```{r Question 1a}
# Define the tsp_distance function to calculate the total distance of a tour
tsp_distance <- function(distances, solution) {
  n <- length(solution)
  total_distance <- 0
  for (i in 1:(n - 1)) {
    total_distance <- total_distance + distances[solution[i], solution[i + 1]]
  }
  total_distance <- total_distance + distances[solution[n], solution[1]]
  return(total_distance)
}

# Define the tsp_2opt_tabu_search function to perform the 2-opt Tabu search
tsp_2opt_tabu_search <- function(distances, max_iterations = 1000, tabu_list_length = 5, max_non_improve = 100) {
  # Get the number of cities
  n <- nrow(distances)
  
  # Initialize the solution with a randomly generated tour
  solution <- sample(1:n)
  
  # Initialize the best solution found so far
  best_solution <- solution
  
  # Initialize the best distance found so far
  best_distance <- tsp_distance(distances, solution)
  
  # Initialize the tabu list
  tabu_list <- rep(-1, tabu_list_length)
  
  # Initialize the number of consecutive non-improving moves
  non_improve <- 0
  
  # Repeat until a stopping criterion is met
  for (iter in 1:max_iterations) {
    # Generate a set of candidate solutions by making 2-opt moves
    candidate_solutions <- list()
    candidate_distances <- numeric()
    idx <- 1
    
    for (i in 1:(n - 1)) {
      for (j in (i + 1):n) {
        candidate <- solution
        candidate[i:j] <- rev(candidate[i:j])
        candidate_solutions[[idx]] <- candidate
        candidate_distances[idx] <- tsp_distance(distances, candidate)
        idx <- idx + 1
      }
    }
    
    # Select the best solution that is not tabu
    feasible_candidates <- setdiff(1:length(candidate_distances), tabu_list)
    best_candidate_idx <- feasible_candidates[which.min(candidate_distances[feasible_candidates])]
    best_candidate_distance <- candidate_distances[best_candidate_idx]
    
    if (best_candidate_distance < best_distance) {
      solution <- candidate_solutions[[best_candidate_idx]]
      best_solution <- solution
      best_distance <- best_candidate_distance
      non_improve <- 0
    } else {
      non_improve <- non_improve + 1
    }
    
    # Update the tabu list
    tabu_list <- c(best_candidate_idx, tabu_list[1:(tabu_list_length - 1)])
    
    # Check the stopping criterion
    if (non_improve >= max_non_improve) {
      break
    }
  }
  
  return(list(solution = best_solution, distance = best_distance))
}

# Define the distance matrix
distances <- matrix(c(
  0, 5, 4, 8, 7, 5,
  5, 0, 4, 6, 3, 2,
  4, 4, 0, 5, 8, 1,
  8, 6, 5, 0, 6, 4,
  7, 3, 8, 6, 0, 6,
  5, 2, 1, 4, 6, 0), nrow = 6, byrow = TRUE)

# Run the algorithm 5 times
for (i in 1:5) {
  result <- tsp_2opt_tabu_search(distances, max_iterations = 1000, tabu_list_length = 5, max_non_improve = 100)
  cat("Run", i, "- Best tour:", result$solution, "Distance:", result$distance, "\n")
}



```

b) Now change the distance matrix to
0, 9, 8, 6, 3, 10, 12
10, 0, 4, 5, 8, 1, 5
4, 6, 0, 5, 6, 4, 6
7, 3, 8, 0, 6, 6, 3
6, 2, 1, 4, 0, 6, 7
4, 7, 9, 10, 5, 0, 6
5, 2, 1, 4, 6, 13, 0
Give the result and best tour. Again, run it a few times to see if the best tour changes.
```{r Question 1b}
# Define the tsp_distance function to calculate the total distance of a tour
tsp_distance <- function(distances, solution) {
  n <- length(solution)
  total_distance <- 0
  for (i in 1:(n - 1)) {
    total_distance <- total_distance + distances[solution[i], solution[i + 1]]
  }
  total_distance <- total_distance + distances[solution[n], solution[1]]
  return(total_distance)
}

# Define the tsp_2opt_tabu_search function to perform the 2-opt Tabu search
tsp_2opt_tabu_search <- function(distances, max_iterations = 1000, tabu_list_length = 5, max_non_improve = 100) {
  # Get the number of cities
  n <- nrow(distances)
  
  # Initialize the solution with a randomly generated tour
  solution <- sample(1:n)
  
  # Initialize the best solution found so far
  best_solution <- solution
  
  # Initialize the best distance found so far
  best_distance <- tsp_distance(distances, solution)
  
  # Initialize the tabu list
  tabu_list <- rep(-1, tabu_list_length)
  
  # Initialize the number of consecutive non-improving moves
  non_improve <- 0
  
  # Repeat until a stopping criterion is met
  for (iter in 1:max_iterations) {
    # Generate a set of candidate solutions by making 2-opt moves
    candidate_solutions <- list()
    candidate_distances <- numeric()
    idx <- 1
    
    for (i in 1:(n - 1)) {
      for (j in (i + 1):n) {
        candidate <- solution
        candidate[i:j] <- rev(candidate[i:j])
        candidate_solutions[[idx]] <- candidate
        candidate_distances[idx] <- tsp_distance(distances, candidate)
        idx <- idx + 1
      }
    }
    
    # Select the best solution that is not tabu
    feasible_candidates <- setdiff(1:length(candidate_distances), tabu_list)
    best_candidate_idx <- feasible_candidates[which.min(candidate_distances[feasible_candidates])]
    best_candidate_distance <- candidate_distances[best_candidate_idx]
    
    if (best_candidate_distance < best_distance) {
      solution <- candidate_solutions[[best_candidate_idx]]
      best_solution <- solution
      best_distance <- best_candidate_distance
      non_improve <- 0
    } else {
      non_improve <- non_improve + 1
    }
    
    # Update the tabu list
    tabu_list <- c(best_candidate_idx, tabu_list[1:(tabu_list_length - 1)])
    
    # Check the stopping criterion
    if (non_improve >= max_non_improve) {
      break
    }
  }
  
  return(list(solution = best_solution, distance = best_distance))
}

# Define the distance matrix
distances <- matrix(c(
0, 9, 8, 6, 3, 10, 12,
10, 0, 4, 5, 8, 1, 5,
4, 6, 0, 5, 6, 4, 6,
7, 3, 8, 0, 6, 6, 3,
6, 2, 1, 4, 0, 6, 7,
4, 7, 9, 10, 5, 0, 6,
5, 2, 1, 4, 6, 13, 0), nrow = 7, byrow = TRUE)

# Run the algorithm 5 times
for (i in 1:5) {
  result <- tsp_2opt_tabu_search(distances, max_iterations = 1000, tabu_list_length = 5, max_non_improve = 100)
  cat("Run", i, "- Best tour:", result$solution, "Distance:", result$distance, "\n")
}


```
a) Run the simulated annealing algorithm for the Rosenbrock function and look at the output.
It is trying to find a minimum for the (complicated) Rosenbrock function, which is used
to test search algorithms. What is the expression for this function? What is the minimum
found? Change the Rosenbrock function by changing the first x^2_1
to just x_1. What is the new minimum found? Give your output.
```{r Question 2a}
# Load necessary libraries
library(stats)

##### Rosenbrock function
# Minimum at f(1,1) = 0
ro <- function(x){
  100 * (x[2] - x[1]^2)^2 + (1 - x[1])^2
}

# Store the function values at each iteration
trace <- list()

# Custom wrapper to store function values
ro_trace <- function(x) {
  val <- ro(x)
  trace <<- append(trace, list(list(par = x, value = val)))
  return(val)
}

# Random start values. Example arguments for the relatively simple Rosenbrock function.
start_values <- runif(2, min = -1, max = 1)

# Perform optimization using BFGS method with box constraints
ro_sa <- optim(
  par = start_values,
  fn = ro_trace,
  method = "L-BFGS-B",  # Limited-memory Broyden-Fletcher-Goldfarb-Shanno with Box constraints
  lower = c(-5, -5),
  upper = c(5, 5),
  control = list(trace = 1, maxit = 550)
)

# Extract function values from trace
func_values <- sapply(trace, function(x) x$value)
par_values <- t(sapply(trace, function(x) x$par))

# Visual inspection of function values over iterations
plot(func_values, type = "l", main = "Function Value Over Iterations", xlab = "Iteration", ylab = "Function Value")

# If you want to visualize the parameter values over iterations
par(mfrow = c(2, 1)) # 2 plots in one column
plot(par_values[,1], type = "l", main = "Parameter x[1] Over Iterations", xlab = "Iteration", ylab = "x[1]")
plot(par_values[,2], type = "l", main = "Parameter x[2] Over Iterations", xlab = "Iteration", ylab = "x[2]")


```
b) Run the simulated annealing algorithm for the Holder function and look at the output.
It is trying to find a minimum for the (complicated) Holder function, which is also a
test function for search algorithms. What is the expression for this function? What is
the minimum found? Change the Rosenbrock function by changing the first factor to
2 ∗ sin(x) ∗ cos(x). What is the new minimum found? Give your output.
```{r Question 2b}
library(optimization)

##### Holder table function
# 4 minima at
# f(8.055, 9.665) = -19.2085
# f(-8.055, 9.665) = -19.2085
# f(8.055, -9.665) = -19.2085
# f(-8.055, -9.665) = -19.2085
ho <- function(x){
  x1 <- x[1]
  x2 <- x[2]
  fact1 <- sin(x1) * cos(x2)
  fact2 <- exp(abs(1 - sqrt(x1^2 + x2^2) / pi))
  y <- -abs(fact1 * fact2)
  return(y)
}

# Non-random start values. Example arguments for the
# relatively complex Holder table function.
ho_sa <- optim_sa(fun = ho,
                  start = c(1, 1),
                  lower = c(-10, -10),
                  upper = c(10, 10),
                  trace = TRUE,
                  control = list(dyn_rf = FALSE,
                                 rf = 1.6,
                                 t0 = 10,
                                 nlimit = 200,
                                 r = 0.6,
                                 t_min = 0.1
                  )
)

# Visual inspection.
plot(ho_sa)
# Plot just gives the function values at increasing iterations
plot(ho_sa, type = "contour")
# Specifying type as contour gives you a nice picture of
# the function value landscape with contour lines and shows
# where the iterations go on the landscape


```

```{r Question 2c}

```
a) Run the genetic algorithm program for the knapsack problem. The given maximum weight
here is 850. Change it to 900 and give the new output.
```{r Question 3a}
library("GA")
require("GA")
p <- c(120, 139, 149, 150, 156, 163, 173, 184, 192, 201, 210, 214, 221, 229, 240)
w <- c(72, 73, 77, 80, 82, 87, 90, 94, 98, 106, 110, 113, 115, 118, 120)
W <- 850
knapsack <- function(x) {f <- sum(x * p)
penalty <- sum(w) * abs(sum(x * w)-W)
f - penalty
}
GA <- ga(type = "binary", fitness = knapsack, nBits = length(w), pmutation = 0.05,
maxiter = 1000, elitism=2, run = 200, popSize = 15)
summary(GA)


p <- c(120, 139, 149, 150, 156, 163, 173, 184, 192, 201, 210, 214, 221, 229, 240)
w <- c(72, 73, 77, 80, 82, 87, 90, 94, 98, 106, 110, 113, 115, 118, 120)
W <- 900
knapsack <- function(x) {f <- sum(x * p)
penalty <- sum(w) * abs(sum(x * w)-W)
f - penalty
}
GA <- ga(type = "binary", fitness = knapsack, nBits = length(w), pmutation = 0.05,
maxiter = 1000, elitism=2, run = 200, popSize = 15)
summary(GA)
```
b) Run the genetic algorithm for the sin function. Change the interval to (−5,5). Give the
new maximum.
```{r Question 3b}

library("GA")
f <- function(x) 2*sin(x)*(x^3 - x^2)
#students can change this function to anything
min <- -5
max <- +5
curve(f, min, max)
fitness <- function(x) f(x)
GA <- ga(type = "real-valued",
fitness = fitness, lower=min, upper = max)
plot(GA)

curve (f, min, max)
points(GA@solution, GA@fitnessValue, col = "blue", pch = 20)
legend(x = -1, y = -20, legend = "global optimum", title = "Best:", pch = 10,col = "blue")

```

```{r Question 3cqla}
library("GA")
library("datasets")

# Define the distance matrix
TPSM <- matrix(c(
  0, 5, 4, 8, 7, 5,
  5, 0, 4, 6, 3, 2,
  4, 4, 0, 5, 8, 1,
  8, 6, 5, 0, 6, 4,
  7, 3, 8, 6, 0, 6,
  5, 2, 1, 4, 6, 0
), nrow = 6, byrow = TRUE)

# Check dimensions of TPSM
print(dim(TPSM))

# Define tspFitness function
tspFitness <- function(tour, distMatrix) {
  tour <- c(tour, tour[1])
  route <- embed(tour, 2)[,2:1]
  sum(distMatrix[route])
}

# Define GA parameters
GA_params <- list(type = "permutation", 
                  fitness = tspFitness, 
                  distMatrix = as.matrix(TPSM),
                  lower = 1, 
                  upper = nrow(TPSM),  # Use nrow(TPSM) as the upper bound
                  popSize = 50, 
                  maxiter = 5000,
                  run = 1,  # Run the GA only once
                  pmutation = 0.2)

# Run the GA algorithm multiple times
best_tours <- list()
best_distances <- numeric()

for (i in 1:5) {
  GA <- do.call(ga, GA_params)  # Using do.call to pass GA_params
  best_tours[[i]] <- GA@solution
  best_distances[i] <- GA@fitnessValue
}

# Print results
for (i in 1:5) {
  cat("Run", i, "- Best Tour:", best_tours[[i]], "Distance:", best_distances[i], "\n")
}

```

```{r Question 3c q1b}
library("GA")
library("datasets")

# Define the distance matrix
TPSM <- matrix(c(
0, 9, 8, 6, 3, 10, 12,
10, 0, 4, 5, 8, 1, 5,
4, 6, 0, 5, 6, 4, 6,
7, 3, 8, 0, 6, 6, 3,
6, 2, 1, 4, 0, 6, 7,
4, 7, 9, 10, 5, 0, 6,
5, 2, 1, 4, 6, 13, 0
), nrow = 7, byrow = TRUE)

# Check dimensions of TPSM
print(dim(TPSM))

# Define tspFitness function
tspFitness <- function(tour, distMatrix) {
  tour <- c(tour, tour[1])
  route <- embed(tour, 2)[,2:1]
  sum(distMatrix[route])
}

# Define GA parameters
GA_params <- list(type = "permutation", 
                  fitness = tspFitness, 
                  distMatrix = as.matrix(TPSM),
                  lower = 1, 
                  upper = nrow(TPSM),  # Use nrow(TPSM) as the upper bound
                  popSize = 50, 
                  maxiter = 5000,
                  run = 1,  # Run the GA only once
                  pmutation = 0.2)

# Run the GA algorithm multiple times
best_tours <- list()
best_distances <- numeric()

for (i in 1:5) {
  GA <- do.call(ga, GA_params)  # Using do.call to pass GA_params
  best_tours[[i]] <- GA@solution
  best_distances[i] <- GA@fitnessValue
}

# Print results
for (i in 1:5) {
  cat("Run", i, "- Best Tour:", best_tours[[i]], "Distance:", best_distances[i], "\n")
}

```
